Loop-Aware Optimizations in PyPy's Tracing JIT

### 2011-06-27 talk/iwtc11/paper.tex
get latest number into the table
some clarifications

\authorinfo{H\aa kan Ardö}
           {Centre for Mathematical Sciences, Lund University}
           {hakan@debian.org}
\authorinfo{Carl Friedrich Bolz}
           {Heinrich-Heine-Universität Düsseldorf}
           {cfbolz@gmx.de}
\authorinfo{Maciej Fijałkowski}
           {Unaffiliated}
           {fijall@gmail.com}

@@

Several benchmarks, with few guard failures, executed on the
PyPy Python JIT show over 2
times increase in speed when loop peeling was introduced. This makes
some of them almost match optimized C performance and become over XXX
times faster than CPython.
\end{abstract}

@@

Because $i_0$ is loop-invariant, the addition could be moved out of the loop.
However, we want to get this effect using our existing optimization passes
without changing them too much. To achieve this, we peel one iteration off the
loop before running the optimizations. This peeling gives the following trace:

@@

This scheme is quite powerful and generalizes to other optimizations than just
common subexpression elimination. It allows simple linear optimization passes to
perform loop-aware optimizations, such as loop-invariant code motion without
changing them at all. All that is needed is to peel off one iteration, then
apply simple one-pass optimizations and make sure that the necessary extra
arguments are inserted into the label of the loop itself and the jumps
afterwards. Giving the optimizations two iterations together
gives the optimization enough context to remove operations from the peeled loop,
because it detects that the operation was performed in the preamble already.

@@

\begin{figure}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
 & CPython & Psyco & PyPy  & PyPy & GCC \\
 &         &       & no LP &      & -O3 \\
\hline
conv3(1e5) & 77.89 & 9.52 & 1.77  & 0.68 &  0.59 \\
\hline
conv3(1e6) & 77.15 & 9.58 & 1.69  & 0.77 &  0.74 \\
\hline
conv3x3(1000) & 236.96 & 128.88 & 0.70  & 0.41 &  0.25 \\
\hline
conv3x3(3) & 23.85 & 12.77 & 0.10  & 0.07  &  0.27 \\
\hline
conv5(1e5) & 122.54 & 16.67 & 1.86  & 1.05 &  0.65\\
\hline
conv5(1e6) & 125.77 & 16.80 & 1.92 & 1.09 &  0.80 \\
\hline
dilate3x3(1000) & 23.29 & 12.99 & 0.41  & 0.39 &  0.26 \\
\hline
sobel(1000) & 181.49 & 95.05 & 0.71 & 0.42 & 0.20 \\
\hline
sqrt(Fix16) & 744.35 & 421.65 & 3.93  & 2.14  & 0.96 \\
\hline
sqrt(float) & 24.21 & 5.52 & 1.36 & 1.00 & 0.98\\
\hline
sqrt(int) & 20.84 & 1.78 & 2.26 & 1.82  & 0.80 \\
\hline
\hline
Variations & - & - & $\pm 0.03$ & $\pm 0.01$ & $\pm 0.01$ \\
\hline
\end{tabular}
}
\end{center}
\label{fig:benchmarks}
\caption{Benchmark Results in Seconds. Arrays of length $10^5$ and
  $10^6$ and matrixes of size $1000\times 1000$ and $1000000 \times
  3$ are used. The one used in each benchmark is indicated in
  the leftmost column. For the matrixes, only the number of rows are
  specified.} 
\end{figure}

@@

Mike Pall, the author of LuaJIT\footnote{\texttt{http://luajit.org/}} seems to
have developped the described technique independently. There are no papers about
LuaJIT but the author of it writes on a mailing list: "The LOOP pass does
synthetic unrolling of the recorded IR, combining copy-substitution with
redundancy elimination to achieve code hoisting. The unrolled and
copy-substituted instructions are simply fed back into the compiler pipeline,
which allows reuse of all optimizations for redundancy elimination. Loop
recurrences are detected on-the-fly and a minimized set of PHIs is generated."
\cite{pall_luajit_2009}

>>>>

\authorinfo{H\aa kan Ardö}
           {Centre for Mathematical Sciences, Lund University}
           {hakan@debian.org}
\authorinfo{Carl Friedrich Bolz}
           {Heinrich-Heine-Universität Düsseldorf}
           {cfbolz@gmx.de}
\authorinfo{Maciej Fijałkowski}
           {}
           {fijall@gmail.com}

@@

Several benchmarks, with few guard failures, executed on the
PyPy Python JIT show over 2
times increase in speed when loop peeling was introduced. This makes
some of them almost match optimized C performance and become over 900
times faster than CPython.
\end{abstract}

@@

Because $i_0$ is loop-invariant, the addition could be moved out of the loop.
However, we want to get this effect using our existing optimization passes
without changing them too much. Simple optimizations with one forward pass
cannot directly get this effect: They just look at the trace without taking
into account that the trace executes many times in a row. Therefore to achieve
loop-invariant code motion, we peel one iteration off the loop before running
the optimizations. This peeling gives the following trace:

@@

This scheme is quite powerful and generalizes to other optimizations than just
common subexpression elimination. It allows simple linear optimization passes to
perform loop-aware optimizations, such as loop-invariant code motion without
changing them at all. All that is needed is to peel off one iteration, then
apply simple one-pass optimizations and make sure that the necessary extra
arguments are inserted into the label of the loop itself and the jumps
afterwards.

This is the key insight of the proposed implementation scheme: Giving an
optimization two iterations together at the same time gives the optimization
enough context to remove operations from the peeled loop, because it detects
that the operation was performed in the preamble already. Thus at runtime these
moved operations are only executed once when entering the loop and the results
are reused in further iterations.

@@

\begin{figure}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
 & CPython & Psyco & PyPy  & PyPy & GCC \\
 &         &       & no LP &      & -O3 \\
\hline
conv3(1e5) & 77.89 & 9.52 & 1.77  & 0.68 &  0.59 \\
\hline
conv3(1e6) & 77.15 & 9.58 & 1.69  & 0.77 &  0.74 \\
\hline
conv3x3(1000) & 233.54 & 125.40 & 0.57 & 0.27 & 0.25 \\
\hline
conv3x3(3) & 234.45 & 126.28 & 0.60 & 0.31 & 0.28 \\
\hline
conv5(1e5) & 122.54 & 16.67 & 1.86  & 1.05 &  0.65\\
\hline
conv5(1e6) & 125.77 & 16.80 & 1.92 & 1.09 &  0.80 \\
\hline
dilate3x3(1000) & 232.51 & 125.85 & 3.89 & 3.69 & 0.25 \\
\hline
sobel(1000) & 181.49 & 95.05 & 0.71 & 0.42 & 0.20 \\
\hline
sqrt(Fix16) & 744.35 & 421.65 & 3.93  & 2.14  & 0.96 \\
\hline
sqrt(float) & 24.21 & 5.52 & 1.36 & 1.00 & 0.98\\
\hline
sqrt(int) & 20.84 & 1.78 & 2.26 & 1.82  & 0.80 \\
\hline
\hline
Variations & - & - & $\pm 0.03$ & $\pm 0.01$ & $\pm 0.01$ \\
\hline
\end{tabular}
}
\end{center}
\label{fig:benchmarks}
\caption{Benchmark Results in Seconds. Arrays of length $10^5$ and
  $10^6$ and matrixes of size $1000\times 1000$ and $1000000 \times
  3$ are used. The one used in each benchmark is indicated in
  the leftmost column. For the matrixes, only the number of rows are
  specified.} 
\end{figure}

@@

Mike Pall, the author of LuaJIT\footnote{\texttt{http://luajit.org/}} seems to
have developped the described technique independently. There are no papers about
LuaJIT but the author of it writes on a mailing list: ``The LOOP pass does
synthetic unrolling of the recorded IR, combining copy-substitution with
redundancy elimination to achieve code hoisting. The unrolled and
copy-substituted instructions are simply fed back into the compiler pipeline,
which allows reuse of all optimizations for redundancy elimination. Loop
recurrences are detected on-the-fly and a minimized set of PHIs is generated.''
\cite{pall_luajit_2009}
